\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-11 jasa/paper-jasa.tex
update minimum for new figures

As a simple and illustrative example, let us first consider the approximation of
the log-likelihood ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\gamma) &= (1-\gamma)\frac{p_{c_0}(\mathbf{x}) +  p_{c_1}(\mathbf{x})}{2}   + \gamma \, p_{c_2}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
figure and will be used later for inference.

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{Neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{Random forest}
        \label{fig:1d}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less accurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5a.pdf}
        \caption{  }
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5b.pdf}
        \caption{ }
        \label{fig:5b}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5c.pdf}
        \caption{ }
        \label{fig:5c}
    \end{subfigure}
    \caption{Inference from exact and approximate likelihood ratios. The red dot corresponds
    to the true values $(\alpha=1, \beta=-1)$ used to generate $\mathcal{D}$, the green dot is the MLE
    $(\hat \alpha=0.896, \hat \beta=-0.984)$ from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=0.893, \hat \beta=-1.135)$ from the approximate likelihood.
    1, 2 and 3-$\sigma$ contours are shown in white.
             (\ref{fig:5a}) The exact log-likelihood for the observed data ${\cal D}$.
             (\ref{fig:5b}) The approximate log-likelihood  evaluated on a coarse $15 \times 15$ fixed grid.
             (\ref{fig:5c}) A Gaussian Process surrogate of the approximate log-likelihood ratio estimated from a Bayesian optimization procedure. White dots show the  parameter points sampled during the optimization process.}
    \label{fig:5}
\end{figure}

Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.
We use the objective function
\begin{equation}\label{eqn:obj-ex2}
    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
\end{equation}
that we approximate by exploiting the fact that
\begin{equation}\label{eqn:obj-ex2-expanded}
\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
\end{equation}
Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
parameterized approximated  likelihood ratio, in this case based on a parameterized
2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
Since the generative model is not expensive, the classifier output is calibrated
on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
Minimizing the constructed approximation of the objective with
Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=0.893, \hat \beta=-1.135$). For
comparison, Figure~\ref{fig:5b} shows the objective value as directly
approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
is a solution to this issue, but is left for now as an area for future work.
Overall, this example further illustrates and confirms the ability
of the proposed method for inference, even in the case of multi-dimensional data where
direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
are often difficult to construct.

@@

In this setting, a classifier is usually built from a large sample of data
$\{\mathbf{x}_i, y_i\}$ generated with some nominal values of the parameters
$\nu=\nu_0$, with $y=0$ for samples drawn from the background component and
$y=1$ for samples drawn from the signal component. Importantly, the $y=1$ label
corresponds to the signal only hypothesis, and not to the alternate
signal-plus-background hypothesis. The resulting classifier approximates the
regression function
$\frac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
one to one with the likelihood ratio of the null to the alternate
$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
$y=1$ label to the signal component has advantages because it helps the
classifier focus its capacity on the relevant regions in the feature space,
particularly when the signal is a very small perturbation to the background
(i.e. $\mu \ll 1$). Once the classifier is trained, large samples of data are
generated from $p_s(\mathbf{x} | \nu)$ and $p_b(\mathbf{x} | \nu)$ and we
estimate the distributions $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ and
$\hat{p}_b(\hat s(\mathbf{x}) | \nu)$ continuously parameterized in $\nu$.

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two
densities from finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.

>>>>

As a simple and illustrative example, let us first consider the approximation of
the log-likelihood ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\gamma) &= (1-\gamma)\frac{p_{c_0}(\mathbf{x}) +  p_{c_1}(\mathbf{x})}{2}   + \gamma \, p_{c_2}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.25^2)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
figure and will be used later for inference.

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{Neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{Random forest}
        \label{fig:1d}
    \end{subfigure}

    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less accurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5a.pdf}
        \caption{  }
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5b.pdf}
        \caption{ }
        \label{fig:5b}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5c.pdf}
        \caption{ }
        \label{fig:5c}
    \end{subfigure}
    \caption{Inference from exact and approximate likelihood ratios. The red dot corresponds
    to the true values $(\alpha=1, \beta=-1)$ used to generate $\mathcal{D}$, the green dot is the MLE
    $(\hat \alpha=1.012, \hat \beta=-0.9221)$  from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=1.008, \hat \beta=-1.004)$ from the approximate likelihood.
    1, 2 and 3-$\sigma$ contours are shown in white.
             (\ref{fig:5a}) The exact log-likelihood for the observed data ${\cal D}$.
             (\ref{fig:5b}) The approximate log-likelihood  evaluated on a coarse $15 \times 15$ fixed grid.
             (\ref{fig:5c}) A Gaussian Process surrogate of the approximate log-likelihood ratio estimated from a Bayesian optimization procedure. White dots show the  parameter points sampled during the optimization process.}
    \label{fig:5}
\end{figure}

Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.
We use the objective function
\begin{equation}\label{eqn:obj-ex2}
    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
\end{equation}
that we approximate by exploiting the fact that
\begin{equation}\label{eqn:obj-ex2-expanded}
\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
\end{equation}
Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
parameterized approximated  likelihood ratio, in this case based on a parameterized
2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
Since the generative model is not expensive, the classifier output is calibrated
on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
Minimizing the constructed approximation of the objective with
Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. For
comparison, Figure~\ref{fig:5b} shows the objective value as directly
approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
is a solution to this issue, but is left for now as an area for future work.
Overall, this example further illustrates and confirms the ability
of the proposed method for inference, even in the case of multi-dimensional data where
direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
are often difficult to construct.

@@

In this setting, a classifier is usually built from a large sample of data
$\{\mathbf{x}_i, y_i\}$ generated with some nominal values of the parameters
$\nu=\nu_0$, with $y=0$ for samples drawn from the background component and
$y=1$ for samples drawn from the signal component. Importantly, the $y=1$ label
corresponds to the signal only hypothesis, and not to the alternate
signal-plus-background hypothesis. The resulting classifier approximates the
regression function
$\sfrac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
one to one with the likelihood ratio of the null to the alternate
$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
$y=1$ label to the signal component has advantages because it helps the
classifier focus its capacity on the relevant regions in the feature space,
particularly when the signal is a very small perturbation to the background
(i.e. $\mu \ll 1$). Once the classifier is trained, large samples of data are
generated from $p_s(\mathbf{x} | \nu)$ and $p_b(\mathbf{x} | \nu)$ and we
estimate the distributions $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ and
$\hat{p}_b(\hat s(\mathbf{x}) | \nu)$ continuously parameterized in $\nu$.

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two
densities from finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1)$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.
